---
title: "Who Talks to Whom?"
description: |
  A few visualizations based on the `speaker` and `addressee` columns.
author:
  - name: Mathieu Glachant
    url: {}
date: 2022-10-11
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
---

Our `said.csv` data file contains information about who said a line of dialog, and to whom. 

What can we do with this data for the chapters edited to date, i.e. the first twelve?

```{r setup}
library(tidyverse)
library(gt)

PATH_TO_SAID <- "../../data/said.csv"

said <- read.csv(PATH_TO_SAID) %>%
  filter(
    book == 1,
    chapter <= 12
  )
```

## Who Speaks? Who Listens?

### Top Speakers

Let's build a list of characters with a speaking part.

We need to:

  - remove the asides to the reader and their rants to Mycroft,
  - eliminate duplicate rows that are not relevant, and 
  - rank them by number of lines of dialog.

```{r top_speakers}
top_speakers <- said  %>%
  # Remove the reader from either side of the conversation
  filter(
    speaker != "#reader",
    addressee != "#reader",
  ) %>%
  # Group by line AND by speaker to deduplicate lines
  group_by(line, speaker) %>%
  summarize(.groups = "drop") %>%
  # Group by speaker and count
  group_by(character = speaker) %>%
  summarize(speaks = n()) %>%
  # Arrange by count
  arrange(desc(speaks))
```

Here are the top five out of `r nrow(top_speakers)` speakers:

```{r show_top_speakers, echo=FALSE}
top_speakers %>%
  head(5) %>%
  gt()
```

### Top Addressees

We can do the same for characters who are directly addressed.

```{r top_addressees}
top_addressees <- said  %>%
  # Remove the reader from either side of the conversation
  filter(
    speaker != "#reader",
    addressee != "#reader",
  ) %>%
  # Group by line AND by addressee to deduplicate lines
  group_by(line, addressee) %>%
  summarize(.groups = "drop") %>%
  # Group by addressee and count
  group_by(character = addressee) %>%
  summarize(spokenTo = n()) %>%
  # Arrange by count
  arrange(desc(spokenTo))
```

Here are the top five out of `r nrow(top_addressees)` addressees:

```{r show_top_addressees, echo=FALSE}
top_addressees %>%
  head(5) %>%
  gt()
```

### Characters Engaged in Dialog

We can now combine the two lists using a join^[If we use a `inner_join()` any characters that are not on both lists will be dropped.] by the `character` column.

```{r characters}
characters <- inner_join(
  top_speakers,
  top_addressees,
  by = "character"
  )
```

Tables are nice and all, but this seems like a good time for a graph. Here are the top 20 out of `r nrow(characters)` characters by lines of dialog:

```{r plot_characters, echo=FALSE}
characters %>%
  head(20) %>%
  gather(speaks, spokenTo, key = "Role", value = lines) %>%
  ggplot(
    mapping = aes(
      y = reorder(character, lines),
      x = lines,
      fill = Role)) +
  geom_bar(position = "stack", stat="identity") +
  labs(
      title = "Top Twenty Characters",
      subtitle = "with a speaking part",
      x = "Lines of dialog",
      y = NULL,
    ) +
  theme(legend.position = c(0.8, 0.25))
```

No real surprise there, Carlyle and Mycroft are our two chatter boxes, even when we remove the reader from the conversation, and it falls off pretty fast after that. Another _Power Law Distribution_?

## Does Anyone Speak More Often Than They Are Spoken To?


